---
title: "lab_11"
author: "sean mussenden"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. You'll need tidyverse, tidycensus, janitor, sf and tigris, and you'll also need to install your census API key.

## Load data

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(sf)
library(tigris)
#census_api_key("9019f6357e35c5c0207c34c0f624606b6449e709")
```

**Task** Create a codeblock and load the WV slice of the PPP loan data, which is in the data folder and is called lab_11.rds.

```{r}
wv_ppp_data <- read_rds("data/lab_11.rds")
```

## Questions 

**Q1.** You are interested in doing some reporting on how West Virginia companies in the "Mining, Quarrying, and Oil and Gas Extraction" industry (https://www.naics.com/six-digit-naics/?code=21) used PPP loans during the pandemic (West Virginia is a big coal mining state).  

Then produce a choropleth map that shows West Virginia counties shaded according to the total amount of loans to that industry per 100,000 population. 

Use the magma color palette, and shade on a "log" scale. Also produce a table that shows the total amount of loans per 100,000 population for each county, ranked from highest to lowest, along with the total amount of loans and the total number of loans. 

Then answer these questions:

* Based on the map, what region of the state appears to have gotten more loans in this industry? What geographic pattern do you notice in the data?
* Based on the table, what county got the highest total loan amount? What was the total amount per 100K population, the total amount of loans (not population adjusted), how many loans were there, and what's the population?

**A1.** Based on the map, I have to say that most of the west in West Virginia is gaining loans from "Mining, Quarrying, and Oil and Gas Extraction" industries. The east part of the state lacks in these loans. Most of the north and west of the state showing that they represent a big coal mining state.
Ritchie county has the most amount per 100k. 
Amount per 100k is 77752742.79.
Total amount of loans is 7653980.00.
Total number of loans is 120.
Population: 9844.


```{r}
mine_oil_wv_ppp <- wv_ppp_data %>%
  filter(naics_code >= "21000", naics_code <= "21999") %>%
  group_by(project_county_name) %>%
  summarise(
    total_loans = sum(amount)
  ) %>%
  arrange(desc(total_loans))
 
```

```{r}
wv_county_population <- get_acs(geography = "county", 
              variables = c(population = "B01001_001"), 
              state = "WV",
              geometry = TRUE)
```

```{r}
wv_county_population <- wv_county_population %>%
  rename(COUNTY = NAME) %>%
  mutate(COUNTY = toupper(str_remove_all(COUNTY,", West Virginia|County"))) %>%
  mutate(COUNTY = str_trim(COUNTY,side="both")) %>%
  rename(population = estimate) %>%
  select(COUNTY, population, geometry)
```

```{r}
wv_county_loans <- wv_county_population %>%
  left_join(mine_oil_wv_ppp, by=c("COUNTY" = "project_county_name")) %>%
  mutate(total_loans = replace_na(total_loans,0)) %>%
  mutate(loans_per_100k = total_loans/population*100000) %>%
  arrange(desc(loans_per_100k))
```

```{r}
ggplot() + 
  geom_sf(data=wv_county_loans, aes(fill=loans_per_100k)) + 
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log") 
```

```{r}
county_totals <- wv_ppp_data %>%
  group_by(project_county_name) %>%
  summarise(
    total_num_loans = n()
  ) %>%
  arrange(desc(total_num_loans))

all_total_wv <- county_totals %>%
  left_join(wv_county_loans, by=c("project_county_name" = "COUNTY"))

all_total_wv %>%
  arrange(desc(loans_per_100k))
```


**Q2.** You want to understand WHY the spatial pattern you saw above exists.  So you grab a shapefile with the latitude and longitude coordinates of every mine permit in West Virginia from the state Department of Environmental Protection. 

Here's the url: https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json. 

(It's in a slightly different format than the .shp file in the lab. It's another type of shapefile called "geojson".  It will work just the same, though, despite that difference. You can read it directly into R without manually downloading it, by putting the url in the st_read() function we learned about in this week's prelab.) 

This data includes both currently operating mines and old ones that have closed.  You want to examine only mines that have a permit expiration date after Sept. 1, 2021. 

Make a map with a point for each one of these mine permits and put a blank map with the outline of West Virginia counties (which you can get from Tigris) behind it. 

What spatial pattern do you see? How does it compare to the spatial pattern in the first map you made?

**A2.** They are actually pretty similar, except the huge difference is in the west. In the first map, most of the west had some data on it, but this map doesn't for the most part. What is similar is the north, south, and east as they all fill up similarly how they did in the first map. 

```{r}
mine_permit_wv <- st_read("https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json")
```

```{r}
after_sept1_mines <- mine_permit_wv %>%
  filter(expiredate >= "2021/09/01")
```

```{r}
counties <- counties() 

glimpse(counties)
```

```{r}
wv_counties <- counties %>%
  filter(STATEFP == "54")
```

```{r}
ggplot() + 
  geom_sf(data=wv_counties) +
  geom_sf(data=after_sept1_mines) + 
  theme_minimal()
```

**Q3.** To confirm the spatial comparison you identified in question 2, let's create a new map that layers the mine location points on top of our PPP choropleth map we created in question 1. What do you notice about the amount of mines in the county that has the highest total loan amount per 100K?

**A3.** 

```{r}
ggplot() +
  geom_sf(data=wv_county_loans, aes(fill=loans_per_100k)) +
  geom_sf(data=after_sept1_mines) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log") 
```

**Q4.** Starting from the original WV ppp dataframe, examine the loans to companies in "Mining, Quarrying, and Oil and Gas Extraction" in the county with the highest total loan amount per 100K (you can use the table you created in question 1 to identify the top county)?
* What single company accounts for the biggest chunk of the total loan amount? 
* What specific industry are they in, based on their 6-digit NAICS code? Look them up on the web. What services do they provide?
* How many jobs did they save, according to the data?
* What's your hypothesis for why a company in their specific industry would need to seek a PPP loan during the pandemic?

**A4.** Biggest chunk- Hall Drilling, LLC
It's an oil and gas company. They provide services like drilling wells and operating gas fields.
387 jobs were saved.
Demand for gas went down, less gas being sold and less amount of demand. Supply was high but no one was in need for gas during the pandemic. 

```{r}
number_4 <- wv_ppp_data %>%
  filter(naics_code >= "21000", naics_code <= "21999")
```
-30-
